---
title: "Bayesian Methods - Assignment 5"
author: "Ryan Durfey"
date: "May 6, 2016"
output: pdf_document
---

```{r, warning=FALSE, message=FALSE}
library(rjags)
library(runjags)
source("//Users/rdurfey/R_misc/BayesianMethods/DBDA2Eprograms/DBDA2E-utilities.R")

# read in data
dat<-read.csv("//Users/rdurfey/R_misc/BayesianMethods/CourseProject/MScA_32014_BayesianMethods_CourseProjectData.csv")
head(dat)
```


# Hierarchical Model
```{r}
# data prep
y=dat$y
s<-as.numeric(dat$sex)
(Ntotal<-length(y))
(Nsubj<-length(unique(s)))
dataList<-list(y=y,s=s,Ntotal=Ntotal,Nsubj=Nsubj)

# model prep
modelString <- "
model {
  for ( i in 1:Ntotal ) {
    y[i] ~ dbern( theta[s[i]] )
  }
  for ( sIdx in 1:Nsubj ) {
    theta[sIdx] ~ dbeta( 100 , 100 ) # N.B.: 2,2 prior; change as appropriate.
  }
}
"
writeLines( modelString , con="TEMPmodel.txt" )

# initialize chains
initsList = function() {
  thetaInit = rep(0,Nsubj)
  for ( sIdx in 1:Nsubj ) { # for each subject
    includeRows = ( s == sIdx ) # identify rows of this group
    yThisSubj = y[includeRows]  # extract data of this group
    resampledY = sample( yThisSubj , replace=TRUE ) # resample
    thetaInit[sIdx] = sum(resampledY)/length(resampledY) 
  }
  thetaInit = 0.001+0.998*thetaInit # keep away from 0,1
  return( list( theta=thetaInit ) )
}

# send the model to JAGS
parameters = c( "theta")     # The parameters to be monitored
adaptSteps = 500             # Number of steps to adapt the samplers
burnInSteps = 500            # Number of steps to burn-in the chains
nChains = 4                  # nChains should be 2 or more for diagnostics 
numSavedSteps<-50000
nIter = ceiling(numSavedSteps / nChains )

# create, initialize, and adapt the model
jagsModel = jags.model( "TEMPmodel.txt" , data=dataList , inits=initsList , 
                        n.chains=nChains , n.adapt=adaptSteps )

# run burn-in
update( jagsModel , n.iter=burnInSteps )

# the main run
codaSamples = coda.samples( jagsModel , variable.names=parameters , n.iter=nIter)
head(codaSamples)

list.samplers(jagsModel)

# analysis 
summary(codaSamples)
plot(codaSamples)
autocorr.plot(codaSamples,ask=F)
effectiveSize(codaSamples)
gelman.diag(codaSamples)
gelman.plot(codaSamples)

#estimated means
matrix(unlist(lapply(codaSamples,function(z) apply(z,2,mean))),ncol=2,byrow = T)

# plot of posterior densities
plot(density(codaSamples[[1]][,1]),xlim=c(0.31,0.41),ylim=c(0,100))
lines(density(codaSamples[[1]][,2]))
lines(density(codaSamples[[2]][,1]),col="red")
lines(density(codaSamples[[2]][,2]),col="red")
lines(density(codaSamples[[3]][,1]),col="blue")
lines(density(codaSamples[[3]][,2]),col="blue")
lines(density(codaSamples[[4]][,1]),col="green")
lines(density(codaSamples[[4]][,2]),col="green")

# HDIs of each chain
(HDIofChains<-lapply(codaSamples,function(z) cbind(Theta1=HDIofMCMC(z[,1]),Theta2=HDIofMCMC(z[,2]))))

# differences between theta1 and theta2
chainDiffs<-lapply(codaSamples,function(z) z[,2]-z[,1])

(leftBounds<-unlist(lapply(chainDiffs,function(z) HDIofMCMC(z,.95)[1])))
head(chainDiffs[[1]])

# plot of left bound of 95% HDI
plot(density(chainDiffs[[1]]))#,xlim=c(-.5,1),ylim=c(0,3),col="black")
lines(density(chainDiffs[[2]]),col="red")
lines(density(chainDiffs[[3]]),col="blue")
lines(density(chainDiffs[[4]]),col="green")
abline(v=leftBounds,col=c("black","red","blue","green"))
```

From the MCMC output and the plots above, we see clear distinction between the two groups. The Male group's distribution is centered around 0.344 while the Female group is centered around 0.384. There is also little to no overlap between the group distributions. This is also evidenced in the differences between the Theta1 and Theta2 not including zero. This means that we can be confident that there is indeed a difference between the groups. Specifically, males seem to have a lower probability of having a postive opinion about Barack Obama than females. However, both genders probabilities are well below fifty percent.



# Separate Models for Each Group
Even though the hierarchical model gave us a pretty solid distinction between the Male and Female groups, we can also try this same example as two separate models where we treat each group's data as separate datasets.
```{r}
# try it again, but with separate samples


# data prep
dat.Male<-dat[s==1,]
dat.Female<-dat[s==2,]
y.Male<-dat.Male$y
y.Female<-dat.Female$y

(Ntotal.Male<-length(y.Male))
(Ntotal.Female<-length(y.Female))

# MLE means
(MLE.meanMale<-sum(y.Male)/Ntotal.Male)
(MLE.meanFemale<-sum(y.Female)/Ntotal.Female)

# create data lists
dataList.Male<-list(y=y.Male,Ntotal=Ntotal.Male)
dataList.Female<-list(y=y.Female,Ntotal=Ntotal.Female)



# create model strings

# modelString.Male
modelString.Male=" 
model {
  for (i in 1:Ntotal) {
    y[i]~dbern(theta)
  }
  theta~dbeta(1,1) #Consider different parameterizations of prior
}
"
writeLines( modelString.Male , con="TempModelMale.txt" )

#ModelString.Contr
modelString.Female=" 
model {
  for (i in 1:Ntotal) {
    y[i]~dbern(theta)
  }
  theta~dbeta(1,1) #Consider different parameterizations of prior
}
"
writeLines( modelString.Female , con="TempModelFemale.txt" )

#Initialize function
initsList = function() {
  thetaInit = rep(0,Nsubj)
  for ( sIdx in 1:Nsubj ) { # for each subject
    includeRows = ( s == sIdx ) # identify rows of this group
    yThisSubj = y[includeRows]  # extract data of this group
    resampledY = sample( yThisSubj , replace=TRUE ) # resample
    thetaInit[sIdx] = sum(resampledY)/length(resampledY) 
  }
  thetaInit = 0.001+0.998*thetaInit # keep away from 0,1
  return( list( theta=thetaInit ) )
}



### Run MCMC for the Male group

#Create model Male
jagsModel.Male<-jags.model(file="TempModelMale.txt",data=dataList.Male,n.chains=4,n.adapt=500)

# Update Male
update(jagsModel.Male,n.iter=600)

# Run Male
codaSamples.Male<-coda.samples(jagsModel.Male,variable.names=c("theta"),n.iter=3334)
list.samplers(jagsModel.Male)

# analyze Male
plot(codaSamples.Male)
autocorr.plot(codaSamples.Male,ask=F)
effectiveSize(codaSamples.Male)
gelman.diag(codaSamples.Male)
gelman.plot(codaSamples.Male)
lapply(codaSamples.Male,mean)

l<-min(unlist(codaSamples.Male))-.005
h<-max(unlist(codaSamples.Male))+.005
histBreaks<-seq(l,h,by=.005)
postHist.Male<-lapply(codaSamples.Male,hist,breaks=histBreaks)

plot(postHist.Male[[1]]$mids,postHist.Male[[1]]$density,type="l",col="black",lwd=2,ylab="Distribution Density (Male)",xlab="Theta",ylim=c(0,80))
lines(postHist.Male[[2]]$mids,postHist.Male[[2]]$density,type="l",col="red",lwd=2)
lines(postHist.Male[[3]]$mids,postHist.Male[[3]]$density,type="l",col="blue",lwd=2)
lines(postHist.Male[[4]]$mids,postHist.Male[[4]]$density,type="l",col="green",lwd=2)
lines(postHist.Male[[3]]$mids,dbeta(postHist.Male[[3]]$mids,1+sum(y.Male),Ntotal.Male-sum(y.Male)+1),type="l",col="purple",lwd=3)
legend("topright",legend=c("Chain1","Chain2","Chain3","Chain4","Theoretical"),col=c("black","red","blue","green"),lwd=2)

(HDIofChains.Contr<-lapply(codaSamples.Male,function(z)
  Theta.Contr=HDIofMCMC(z[,1])))




### Run MCMC for the Female group

#Create model FeFemale
jagsModel.Female<-jags.model(file="TempModelFemale.txt",data=dataList.Female,n.chains=4,n.adapt=500)

# Update Female
update(jagsModel.Female,n.iter=600)

# Run Female
codaSamples.Female<-coda.samples(jagsModel.Female,variable.names=c("theta"),n.iter=3334)
list.samplers(jagsModel.Female)

# analyze Female
plot(codaSamples.Female)
autocorr.plot(codaSamples.Female,ask=F)
effectiveSize(codaSamples.Female)
gelman.diag(codaSamples.Female)
gelman.plot(codaSamples.Female)
lapply(codaSamples.Female,mean)

l<-min(unlist(codaSamples.Female))-.005
h<-max(unlist(codaSamples.Female))+.005
histBreaks<-seq(l,h,by=.005)
postHist.Female<-lapply(codaSamples.Female,hist,breaks=histBreaks)

plot(postHist.Female[[1]]$mids,postHist.Female[[1]]$density,type="l",col="black",lwd=2,ylab="Distribution Density (Female)",xlab="Theta",ylim=c(0,90))
lines(postHist.Female[[2]]$mids,postHist.Female[[2]]$density,type="l",col="red",lwd=2)
lines(postHist.Female[[3]]$mids,postHist.Female[[3]]$density,type="l",col="blue",lwd=2)
lines(postHist.Female[[4]]$mids,postHist.Female[[4]]$density,type="l",col="green",lwd=2)
lines(postHist.Female[[4]]$mids,dbeta(postHist.Female[[4]]$mids,1+sum(y.Female),Ntotal.Female-sum(y.Female)+1),type="l",col="purple",lwd=3)
legend("topright",legend=c("Chain1","Chain2","Chain3","Chain4","Theoretical"),col=c("black","red","blue","green"),lwd=2)

(HDIofChains.Contr<-lapply(codaSamples.Female,function(z)
  Theta.Contr=HDIofMCMC(z[,1])))

#Find differences between Male and Female groups and their HDIs
Male.Female.diffs<-sapply(1:4,function(z) codaSamples.Female[[z]]-codaSamples.Male[[z]])
(HDI.diffs<-apply(Male.Female.diffs,2,function(z) HDIofMCMC(z,.95)))

#Plot densities and the 95% HDI limits
plot(density(Male.Female.diffs[,1]))#,xlim=c(-.5,1),ylim=c(0,3),col="black")
lines(density(Male.Female.diffs[,2]),col="red")
lines(density(Male.Female.diffs[,3]),col="blue")
lines(density(Male.Female.diffs[,4]),col="green")
abline(v=HDI.diffs[1,],col=c("black","red","blue","green"))
```
```{r}
#Compare left bounds of hierarchical model and separate models
rbind(Hierarchical=leftBounds,Separate=HDI.diffs[1,])

```

Just like in the first hierarchical model, we see the same distinction between groups. Males have a lower probability by a few percentage points than females to have a positive opinion about Barack Obama.